State of the Practice for Lattice Boltzmann Method Software

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed Section 5

A small sample of safety and mission-critical applications of LBM includes the
following: designing fuel cells \citep{ZhangEtAl2018}, modelling groundwater
flow \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the
cardiovascular system \citep{SadeghiEtAl2022b, SadeghiEtAl2022,
SadeghiEtAl2020}.  As these examples illustrate, engineers and scientists use
LBM for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety. Given the critical applications of LBM,
users worry about with software qualities like reliability,
robustness, reproducibility, performance, correctness, and verifiability.  Since
their time is valuable, developers of LBM libraries have additional quality
concerns, including maintainability, reusability, and understandability.  With
the quality considerations in mind, and the potentially overwhelming number of
choices for existing LBM libraries, the time is right to assess the state of the
practice. Therefore, we analyze the current state of LBM
software development and compare it to the development practices employed for
research software in general.  We want to highlight success stories that LBM
developers can share amongst their community, and with the broader research
software community, while at the same time looking for areas for potential
future improvement.

@@

As part of filling in the measurement template the LBM packages were examined
for the presence of artifacts, which were then categorized by frequency. We have
grouped the artifacts into common, uncommon, and rare categories in
Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
of the software packages. Uncommon artifacts were found in 8 to 15 (30-63\%) and
rare artifacts were found in 1 to 7 ($<$30\%) of the packages.  We answer
\rqref{RQ_CompareArtifacts} by comparing the artifacts that we observed in LBM
repositories to those observed and recommended for research software in general.
Our comparison may point out areas where some LBM software packages fall short
of current best practices. This is not intended to be a criticism of any
existing packages, especially since in practice not every project needs to
achieve the highest possible quality.  However, rather than delve into the
nuances of which software can justify compromising which practices we will write
our comparison under the ideal assumption that every project has sufficient
resources to match best practices.

@@

In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
row, a checkmark in one of the columns means that the corresponding guideline
recommends this artifact.  The last column shows whether the artifact appears in
the measured set of LBM software, either not at all (blank), commonly (C),
uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
artifact consistently between guidelines and specific MI software, but the
terminology and the contents of artifacts are not standardized.  The challenge
even exists for the ubiquitous README file.  As illustrated by
\citet{PranaEtAl2018}, the content of README files shows significant variation
between projects.  Although some content is reasonably consistent, with 97\% of
README files contain at least one section describing the `What` of the
repository and 88.5 \% offering some `How` content, other categories are more
variable.  For instance, information on `Contribution`, `Why`', and `Who`,
appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively
\citep{PranaEtAl2018}.

The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
popularity of recommending a given artifact, but it does not imply that the most
commonly recommended artifacts are the most important artifacts. Just because an
artifact is not explicitly recommended in a given guideline, does not mean that
the artifact is not valued.  The guideline authors may have excluded it because
it is out of the scope of their recommendations, or outside their experience.
For instance, an artifact related to uninstall is only explicitly mentioned by
\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
value.  They may simply feel that uninstall is implied by install, or they may
have never asked themselves whether separate uninstall instructions are needed.

@@

The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all the
common artifacts, except Palabos does not have theory notes and only three
(ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
system.  The top four packages also have most of the uncommon artifacts. At
the time of data collection, only one of the four packages (Palabos) did not
have a user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. This broken link was later fixed,
but this is not reflected in our data because it was not present at the time of
data collection (mid-2020). Despite the broken link, Palabos does have a
detailed and informative website. As far as we could observe, Ludwig does not
appear to have a publicly visible design and OpenLB does not appear to use a
version control system. However, OpenLB's use of version numbers suggests that
version control may be used privately by the developers.

@@

Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
appear in the software development guidelines: i) list of related publications,
ii) performance information and iii) video guides.  The occurrence of the first
two items in this list are likely more a consequence of how we collected our
list of LBM artifacts, rather than being a unique practice of LBM developers. We
highlighted publications and performance information because we were explicitly
looking for both as part of filling in our measurement template. Related
publications and performance information certainly appear in other research
software projects; the practices just aren't highlighted in guidelines because
they are implicit in other documentation recommendations, such as theory
documentation and user guides. The observation of a video guide artifact is more
novel.  \citet{Fogel2005} recommends videos for open source projects, and other
projects likely use them, but the practice has apparently not yet made it into
research software community's guidelines.  Putting aside the debate of the
effectiveness of learning from print versus video, video documentation is
growing in popularity, with a majority of Generation Z learners preferring video
to print \citep{Genota2018}.

As Table~\ref{Tbl_Guidelines} shows the LBM community participates in most of
the practices we found listed in the general research software guidelines;
however, some recommended practices were not observed or rarely observed. For
instance, API documentation was rarely observed for LBM software, but it is a
frequently recommended artifact \citep{SmithAndRoscoe2018, ThielEtAl2020,
vanGompelEtAl2016, OrvizEtAl2017, SSI2022, Zadka2018}.  In addition to the items
in the last column of Table~\ref{artifactspresent}, we can add the following
community recommended items that we rarely, if ever, observed:

@@

\item \textbf{A roadmap} is considered by some to be a critical document for an
organization to map out the vision and direction for a product offering
\citep{MunchEtAl2019}.  Roadmaps take different forms, but they all aim to
address the following: \begin{inparaenum}[i)]
	\item Where are we now?
	\item Where are we going? and
	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}.
Developers can create a roadmap by the following steps:
\begin{inparaenum}[i)]
	\item define and outline a strategic mission and product vision, 
	\item scan the environment, 
	\item revise and distill the product vision to write the product roadmap, 
	\item estimate the product life cycle and evaluate the mix of planned
development efforts. \end{inparaenum} \citep{VahaniittyEtAl2002}.  A product
roadmap has the benefits of providing continuity of purpose, facilitating
stakeholder collaboration and assisting with prioritization \citep{Pichler2012}.
Although we saw some elements of a roadmap in the LBM projects, we did not
observe a specific artifact devoted to this purpose.

\item \textbf{A code of conduct} documents expectations and standards of ethical
behaviour and explicitly states how developers should treat one another
\citep{TouraniEtAl2017}. A code of conduct outlines rules for communication,
establishes enforcement mechanisms for violations and codifies the spirit of a
community, such that anyone can contribute comfortably regardless of, for
example, ethnicity, gender, or sexual orientation \citep{TouraniEtAl2017}. The
top three most popular code's of conduct from \citet{TouraniEtAl2017} are:
\href{https://ubuntu.com/community/code-of-conduct} {Ubuntu Code of Conduct},
\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
{Contributor Covenant}, \href{https://www.djangoproject.com/conduct/} {Django
Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a code of
conduct on the participation of women. Improved inclusivity means a wider pool
of potential contributors. Moreover, a code of conduct facilitates meeting
ethical obligations for developers that abide by a code of ethics, such as the
IEEE Code of Ethics \citep{IEEE1999}, or the Professional Engineers of Ontario
code of ethics \citep[p.\ 23--24]{PEO2021}. For the LBM software, we did not
observe any codes of conduct, possibly because the number of developers for each
LBM project is small and because the idea hasn't yet occurred to any of the
participants.

>>>>

A small sample of safety and mission-critical applications of LBM includes the
following: designing fuel cells \citep{ZhangEtAl2018}, modelling groundwater
flow \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the
cardiovascular system \citep{SadeghiEtAl2022b, SadeghiEtAl2022,
SadeghiEtAl2020}.  As these examples illustrate, engineers and scientists use
LBM for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety. Given the critical applications of LBM,
users worry about with software qualities like reliability,
robustness, reproducibility, performance, correctness, and verifiability.  Since
their time is valuable, developers of LBM libraries have additional quality
concerns, including maintainability, reusability, and understandability.  With
these quality considerations in mind, and the potentially overwhelming number of choices for existing LBM libraries, the time is right to assess the state of the practice. Therefore, we analyze the current state of LBM
software development and compare it to the development practices employed for
research software in general.  We want to highlight success stories that LBM
developers can share amongst their community, and with the broader research
software community, while at the same time looking for areas for potential
future improvement.

@@

As part of filling in the measurement template the LBM packages were examined
for the presence of artifacts, which were then categorized by frequency. We have
grouped the artifacts into common, uncommon, and rare categories in
Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
of the software packages. Uncommon artifacts were found in 8 to 15 (30-63\%) and
rare artifacts were found in 1 to 7 ($<$30\%) of the packages.  We answer
\rqref{RQ_CompareArtifacts} by comparing the artifacts that we observed in LBM
repositories to those observed and recommended for research software in general.
Our comparison may point out areas where some LBM software packages fall short
of current best practices. This is not intended to be a criticism of any
existing packages, especially since in practice not every project needs to
achieve the highest possible quality.  However, rather than delve into the
nuances of which software can justify compromising which practices, we will write our comparison under the ideal assumption that every project has sufficient resources to match best practices.

@@

In Table~\ref{Tbl_Guidelines}, each row corresponds to an artifact.  For a given
row, a checkmark in one of the columns means that the corresponding guideline
recommends this artifact.  The last column shows whether the artifact appears in
the measured set of LBM software, either not at all (blank), commonly (C),
uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
artifact consistently between guidelines and specific MI software, but the
terminology and the contents of artifacts are not standardized.  The challenge
even exists for the ubiquitous README file.  As illustrated by
\citet{PranaEtAl2018}, the content of README files shows significant variation
between projects.  Although some content is reasonably consistent, with 97\% of
README files containing at least one section describing the `What` of the
repository and 88.5 \% offering some `How` content, other categories are more
variable.  For instance, information on `Contribution`, `Why`, and `Who`,
appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively
\citep{PranaEtAl2018}.

The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
popularity of recommending a given artifact, but it does not imply that the most
commonly recommended artifacts are the most important artifacts. Just because an
artifact is not explicitly recommended in a given guideline does not mean that
the artifact is not valued.  The guideline authors may have excluded it because
it is out of the scope of their recommendations or outside their experience.
For instance, an artifact related to uninstall is only explicitly mentioned by
\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
value.  They may simply feel that uninstall is implied by install, or they may
have never asked themselves whether separate uninstall instructions are needed.

@@

The top four packages (ESPREsSo, Ludwig, Palabos, and OpenLB) have all the
common artifacts, except Palabos does not have theory notes and only three
(ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
system.  The top four packages also have most of the uncommon artifacts. At
the time of data collection, only one of the four packages (Palabos) did not
have a user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. This broken link was later fixed,
but this is not reflected in our data because it was not present at the time of
data collection (mid-2020). Despite the broken link, Palabos does have a
detailed and informative website. As far as we could observe, Ludwig does not
appear to have a publicly visible design and OpenLB does not appear to use a
version control system. However, OpenLB's use of version numbers suggests that
version control may be used privately by the developers.

@@

Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
appear in the software development guidelines: i) list of related publications,
ii) performance information and iii) video guides.  The occurrence of the first
two items in this list are likely more a consequence of how we collected our
list of LBM artifacts, rather than being a unique practice of LBM developers. We
highlighted publications and performance information because we were explicitly
looking for both as part of filling in our measurement template. Related
publications and performance information certainly appear in other research
software projects; the practices just are not highlighted in guidelines because
they are implicit in other documentation recommendations, such as theory
documentation and user guides. The observation of a video guide artifact is more
novel.  \citet{Fogel2005} recommends videos for open source projects, and other
projects likely use them, but the practice has apparently not yet made it into
research software community's guidelines.  Putting aside the debate of the
effectiveness of learning from print versus video, video documentation is
growing in popularity, with a majority of Generation Z learners preferring video
to print \citep{Genota2018}.

As Table~\ref{Tbl_Guidelines} shows, the LBM community participates in most of
the practices we found listed in the general research software guidelines;
however, some recommended practices were not observed or rarely observed. For
instance, API documentation was rarely observed for LBM software, but it is a
frequently recommended artifact \citep{SmithAndRoscoe2018, ThielEtAl2020,
vanGompelEtAl2016, OrvizEtAl2017, SSI2022, Zadka2018}.  In addition to the items
in the last column of Table~\ref{artifactspresent}, we can add the following
community recommended items that we rarely, if ever, observed:

@@

\item \textbf{A roadmap} is considered by some to be a critical document for an
organization to map out the vision and direction for a product offering
\citep{MunchEtAl2019}.  Roadmaps take different forms, but they all aim to
address the following: \begin{inparaenum}[i)]
	\item Where are we now?,
	\item Where are we going?, and
	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}.
Developers can create a roadmap by the following steps:
\begin{inparaenum}[i)]
	\item define and outline a strategic mission and product vision, 
	\item scan the environment, 
	\item revise and distill the product vision to write the product roadmap, and 
	\item estimate the product life cycle and evaluate the mix of planned
development efforts. \end{inparaenum} \citep{VahaniittyEtAl2002}.  A product
roadmap has the benefits of providing continuity of purpose, facilitating
stakeholder collaboration and assisting with prioritization \citep{Pichler2012}.
Although we saw some elements of a roadmap in the LBM projects, we did not
observe a specific artifact devoted to this purpose.

\item \textbf{A code of conduct} documents expectations and standards of ethical
behaviour and explicitly states how developers should treat one another
\citep{TouraniEtAl2017}. A code of conduct outlines rules for communication,
establishes enforcement mechanisms for violations and codifies the spirit of a
community, such that anyone can contribute comfortably regardless of, for
example, ethnicity, gender, or sexual orientation \citep{TouraniEtAl2017}. The
top three most popular code's of conduct from \citet{TouraniEtAl2017} are:
\href{https://ubuntu.com/community/code-of-conduct} {Ubuntu Code of Conduct},
\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
{Contributor Covenant}, and \href{https://www.djangoproject.com/conduct/} {Django
Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a code of
conduct on the participation of women. Improved inclusivity means a wider pool
of potential contributors. Moreover, a code of conduct facilitates meeting
ethical obligations for developers that abide by a code of ethics, such as the
IEEE Code of Ethics \citep{IEEE1999}, or the Professional Engineers of Ontario
code of ethics \citep[p.\ 23--24]{PEO2021}. For the LBM software, we did not
observe any codes of conduct, possibly because the number of developers for each
LBM project is small and because the idea has not yet occurred to any of the
participants.
